Shedding Light on the ``Dark Side'' of Phylogenetic Comparative Methods

### 2015-12-08 Manuscript/response.tex

\textcolor{blue}{Minor Comments
Abstract: This should be in the proper BES-approved format (i.e. as a bulleted list).}

Fixed

\textcolor{blue}{l315-6: A proper citation is needed, I think. Because who wouldn't want to see this cited formally?}

Exactly how a song should be cited is not entirely clear but we have added a the following citation: Manic Street Preachers. 1996. Mr Carbohydrate. A Design for Life [CD 1], track 2. Columbia Records.

@@

\textcolor{blue}{My biggest suggestion is that there can be some tightening of the text in a few places and a better mapping in others to the headings. I’ll detail this specifically below.}

Thanks for the helpful and encouraging comments. :)

\textcolor{blue}{L179 I’d suggest that some of this folklore also comes from within the labs developing PCM. The reason Sam Price and I proposed the R hackathon at NESCent was because each of us had knowledge about PCM issues and solutions that were lab based from the groups where we came from who were developing methods. Some of this folklore is based on good practices, others not but the point is it is being passed down outside of the public literature.}

We agree and make exactly these points on line XX: "Useful PCM folklore is often shared among developers, and among collaborating groups, but is not always shared outside of these circles."

@@

\textcolor{blue}{L224-225 I think you can just say long and tedious or something along those lines. Why is “>20 pages” too long and what gives me a headache may be different to what gives you a “headache”.}

We have removed reference to headaches and also added the point that often equations are necessary.

@@

\textcolor{blue}{L231 “This” is somewhat unclear. I’d make it clear that this review led to you needing to read more than 300 manuscript pages and a book to fully understand the issue.}

We have expended on this to make the point that the difficulty relates to the problem of how much reading is necessary to have sufficient understanding.

@@

\textcolor{blue}{Reviewer 2: Simon Blomberg}

\textcolor{blue}{After reading the abstract of this paper, I was sceptical. However, upon reading it I found that it makes some reasonable points. First I will discuss the negative aspects.}

\textcolor{blue}{My main issue is with sections 2 (beginning line 218) and 3 (beginning line 238). The fact that the literature is too difficult and opaque for some readers is not the fault of the literature. These statistical methods are NOT easy, so researchers will need to do due diligence and take the time to learn this stuff properly (including caveats and assumptions, when they are known and described in the literature). In any statistical analysis, the buck stops with the authors. It is your responsibility to get it right. You can't complain that you are using methods that you don't understand. If that is the case you should go back and learn more about the methods. Complaining that there is too much maths and it gives you a headache (lines 224-225) is not a reason to castigate the authors of methods. Mathematics is the language of modern evolutionary biology and the sooner people get used to that, the better. Don't like maths? Don't do quantitative science. Complaining that the literature is too big and difficult and going ahead with an analysis anyway is just academic laziness. Don't want to get to grips with the literature? Collaborate with someone who knows it, like a statistician. The authors appear to be promoting a new section for such papers, as assumptions and caveats can be in the Intro, Methods, Results and/or Discussion (lines 235-237). Well how about this: read the papers properly! Again, laziness here.}

We have removed reference to headaches and revised these sections substantially. We agree that due diligence is key. However our point is that due diligence is not just the responsibility of readers/users but also of developers. It is not simply laziness on the readers side that leads to misunderstanding. Often papers, especially those with software, are presented as if the method is faultless. This is a symptom of how papers are published but can very easily mislead novice users. Clarity in writing is the responsibility of the write, thorough reading is the responsibility of the reader.

@@

\textcolor{blue}{The authors complain that researchers often jump straight into using a method (line 259-263) without knowing its limitations. Well this is just a problem with due diligence again. It's not the method developer's fault. The authors complain that this seems to be particularly a problem with methods implemented in R. Well, R is free software. You have the source code. If you really want to know how a method works, study the source code! You can't do that for other, proprietary software. You have to trust the company/lab that produces the software that it does what is claimed on the box. Using free software like R is thus more scientific than using proprietary software. It \_allows\_ researchers to do due diligence.}

We agree. However, as we note above we also argue that doing due diligence is a responsibility of both the users (in reading widely and deeply) and the developers (in writing clearly and stating assumptions, limitations etc).

### 2015-12-08 Manuscript/dark.tex
GHTs edits

  \subsection{2. The literature is too technical and/or important details are difficult to locate}
    Although some information is not found in the literature (see above), the majority of assumptions and biases of PCMs are documented somewhere. 
    A big issue for novice methods-users (and often for advanced users too) is that this information can be extremely technical and dense.
    It is not unusual for papers to be full of equations, \textgreater 20 pages long. 
    Such detail and mathematical literacy is critical in describing methods and facilitates testing/assessment, implementation and future developments of the method.
    In most cases equations could not simply be replaced with text.
   However, from the end user perspective it may more helpful to understand the assumptions and caveats.
   Within many comparative methods assumptions and caveats can be found in the Introduction, Methods, Results and/or Discussion of a paper. 
    They are rarely neatly corralled in one place.
    When combined with the vast literature on comparative methods it becomes easy to miss pertinent details. 
     We recently reviewed papers discussing the assumptions and limitations of phylogenetic independent contrasts \citep{felsenstein1985phylogenies}, including how they are related to phylogenetic generalised least squares models \citep{garland2000using,rohle2006comment,blomberg2012independent}. 
    Even with prior knowledge of the key papers and authors to focus on, this resulted in $\approx 300$ manuscript pages and a book to read.
    Methods developers are not responsible for users conducting due diligence or reading the literature. 
    The onus is, quite rightly in our opinion, on the reader/user to ensure that they have a clear understanding of the methods and caveats prior to using them. 
    However, simple steps, such as subheadings that point to caveats and assumptions could add clarity and limit method misuse.

@@

\subsection{How can we solve some of these problems?}
  \subsubsection{1. Simplify, summarise and share}
    Many of the problems above are caused by methods papers being difficult to read. 
    This is not entirely the fault of those writing these papers; most journals have strict word limits so it is often easier to use a one line equation rather than a paragraph of description. 
    As we note above, the onus is on the reader to read extensively and to do due diligence.
    Nonetheless, due diligence is a responsibility at all levels of research and that includes ensuring clarity to all target readers, especially in describing assumptions and limitations.
    One solution is to prepare an accompanying blog post or video explaining the method in less technical terms. 
    Some journals already encourage this (including the British Ecological Society journals; see https://soundcloud.com/besjournals, https://www.youtube.com/user/MethodsEcolEvol), and even if they do not, there is no harm in doing this and hosting it on a personal website. 
    The ability to share ideas with a non-technical audience is a key skill to develop, and may encourage more people to use the method.
    Encouraging increased efforts at reproducibility would also help, by including knitR \citep{Xie:2015aa} reports in Supplementary Material showing exactly how each analysis was run \citep[e.g.][]{fitzjohn2014much}, or by requiring all analyses and code to be available on GitHub or Bitbucket. 
    At the very least, having a list of the main assumptions, biases and caveats of the method somewhere obvious would reduce misuse and provide a place to point people to when they begin using a method (this will be difficult as there are often hidden assumptions in a method, and listing all possible assumptions and caveats may be unfeasible in some cases).

>>>>

  \subsection{2. The literature is too technical and/or important details are difficult to locate}
    Although some information is not found in the literature (see above), the majority of assumptions and biases of PCMs are documented somewhere. 
    A big issue for novice methods-users (and often for advanced users too) is that this information can be extremely technical and dense.
    It is not unusual for papers to be full of equations, \textgreater 20 pages long. 
    Such detail and mathematical literacy is critical in describing methods and facilitates testing/assessment, implementation and future developments of the method.
    In most cases equations could not simply be replaced with text.
   However, from the end user perspective it may more helpful to understand the assumptions and caveats.
   Within many comparative methods assumptions and caveats can be found in the Introduction, Methods, Results and/or Discussion of a paper. 
    They are rarely neatly corralled in one place.
     The sheer volume of literature can become a barrier to understanding, even, and perhaps especially, for the best known methods.
     For example, we recently reviewed papers discussing the assumptions and limitations of phylogenetic independent contrasts \citep{felsenstein1985phylogenies}, including how they are related to phylogenetic generalised least squares models \citep{garland2000using,rohle2006comment,blomberg2012independent}. 
    Even with prior knowledge of the key papers and authors to focus on, this resulted in $\approx 300$ manuscript pages and a book to read.
    The volume of reading itself is perhaps not the key issue, rather it is assessing when you have reached sufficient understanding which may be easy for experienced users but not clear for new users.    
    The combined effects of a vast literature and sometimes opaque assumptions make it easy to miss pertinent details.
    Many of these problems relate to due diligence for both reader and developer. 
    Methods developers are not responsible for making sure that all users read the literature. 
    The onus is, quite rightly in our opinion, on the reader/user to ensure that they have a clear understanding of the methods and caveats prior to using them. 
    However, simple steps, such as subheadings that point to caveats and assumptions, could add clarity and limit method misuse.

@@

\subsection{How can we solve some of these problems?}
  \subsubsection{1. Simplify, summarise and share}
    Many of the problems above are the result of methods that are difficult to communicate in purely descriptive terms.
    This is not entirely the fault of those writing these papers; most journals have strict word limits and a one line equation is generally a more succinct and precise descriptor of a method than a paragraph of text. 
    As we note above, the onus is on the reader to read extensively and to do due diligence.
    Nonetheless, due diligence is a responsibility at all levels of research and that includes ensuring clarity to all target readers, especially in describing assumptions and limitations.
    One solution is to prepare an accompanying blog post or video explaining the method in less technical terms. 
    Some journals already encourage this (including the British Ecological Society journals; see https://soundcloud.com/besjournals, https://www.youtube.com/user/MethodsEcolEvol), and even if they do not, there is no harm in doing this and hosting it on a personal website. 
    Some package developers provide blog updates (e.g. phytools http://blog.phytools.org) that serve this function. 
    The ability to share ideas with a non-technical audience is a key skill to develop, and may encourage more people to use the method.
    Encouraging increased efforts at reproducibility would also help, by including knitR \citep{Xie:2015aa} reports in Supplementary Material showing exactly how each analysis was run \citep[e.g.][]{fitzjohn2014much}, or by requiring all analyses and code to be available on GitHub or Bitbucket. 
    At the very least, having a list of the main assumptions, biases and caveats of the method somewhere obvious would reduce misuse and provide a place to point people to when they begin using a method (this will be difficult as there are often hidden assumptions in a method, and listing all possible assumptions and caveats may be unfeasible in some cases).
    An important point here is that we do not see these possible solutions as entirely the responsibility of the method developer or package writer. 
    The comparative methods community can help too and indeed often does via platforms such as r-sig-phylo.
    However, the ability to find helpful replies to lists depends on the subject headings and details in the text.
    A longer lasting solution would be to make published methods papers, and especially those that have a software focus, more readily updatable. 
    It should not be necessary to publish a new paper for updates of method.
    One example that gets around this problem is PLoS Currents where published papers can be updated relatively quickly with the full history of updated articles versioned.
